Ages of Planets and Stars Observed by Kepler

### 2024-04-07 paper/ms.tex
ms: clarify quality flag syntax

Can stellar age measurements provide evidence for or against
these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While an apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  Recent studies have
nonetheless found that when accounting for completeness, there is
$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
\citep{2024arXiv240303261V} stars than in the older Kepler sample
\citep[e.g.][]{2020AJ....159..248K}.
%TODO: more kepler citations...

@@

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.
In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Many of
these reported detections have lower variability amplitudes, and
longer periods, than those reported by
\citetalias{McQuillan_2014}.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.  Reviewing the resulting period detections, we
learned that 32 KOIs with orbital and rotation periods within
$\approx$20\% were excluded from the \citetalias{Santos_2019} and
\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
to include these objects.\footnote{In general, excluding stars in this
category would impose a bias against young close-in planets, due to
the natural commensurability between the rotation rates of young
stars, and Kepler's increased sensitivity to short-period objects.
While some of these KOIs are known false positives, our analysis
naturally excludes such objects at later steps.  As one motivating
example, this subset of stars happens to include {\it
e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
based on membership in RSG-5 \citep{Bouma_2022}.  }
The final tally of stars from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our corrected KOI list is
\nuniqstarsantosrot\ Kepler stars with rotation periods.

@@

\subsection{Star Filters}
\label{subsec:filters}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of filters and quality flags, and apply
them in the sub-analyses that follow.  We defined these filters as
follows, each of which is included in Tables~\ref{tab:stars}
and~\ref{tab:planets}.

@@

{\it Final calibration sample}---{\bf todo: mention any extra flags
from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
amplitude cuts you might want to apply to the rotation period catalog}
%The combination of the filters
%described above yields the set of stars that show no evidence for
%binarity or crowding.  However, some of the rotation period analyses
%in Table~\ref{tab:clusters} include additional relevant quality flags.
%For instance, light curves showing multiple photometric periods can
%indicate unresolved binarity.  We used all relevant filters available
%from the original authors if they were designed to select single stars
%with reliable rotation periods.  The final combination of these
%filters with our own flag for possible binarity yields our sample of
%benchmark rotators.

\subsection{Planet Filters}
For the subset of our analysis that focuses on the properties of
planets, we use the following additional filters.

@@

The rotation periods for the stellar and planetary samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods yielded
\nuniqstarsantosrotgyroappl.  Further requiring these rotating stars
to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
over which \texttt{gyro-interp} returns finite age posteriors,
yields \nuniqstarsantosrotteffcut\ stars.
Further requiring them to be apparently single, near the main
sequence, with $\log g$$>$4.2, RUWE$<$1.4, and not flagged as
eclipsing binaries yielded \nuniqstarsantosrotgyroappl\ rotators
amenable for gyrochronology.

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Applying all the same stellar cuts, and
requiring these confirmed and candidate planets to have MES$>$10,
yields \nplwgyroage\ planets with rotation-based ages orbiting
\nplhoststarwgyroage\ stars.  If we additionally require the planets
to be non-grazing ($b<0.8$) based on the Markov Chain Monte Carlo
(MCMC) fits from \citet{Thompson_2018}, we are left with
\nplwgyroagenograzing\ and \nplhoststarwgyroagenograzing\ planets and
stars, respectively.

@@

To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ feature is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Age distribution of stars and planet-hosts in the Kepler field
    inferred from rotation periods}.
    A histogram of samples from the age posteriors for all Kepler
    stars with reported rotation periods and temperatures is shown in
    light gray;
    the age calculation was actually valid for stars that meet the criteria
    discussed in Section~\ref{subsec:filters} (dark blue).
		\label{fig:hist_tgyro}
	}
\end{figure*}

>>>>

Can stellar age measurements provide evidence for or against
these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While an apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
nonetheless found that when accounting for completeness, there is
$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
\citep{2024arXiv240303261V} stars than in the older Kepler sample
\citep[e.g.][]{2020AJ....159..248K}.
%TODO: more kepler citations...

@@

We choose to adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provides a detailed comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.
In addition,
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.  Many of
these reported detections have lower variability amplitudes, and
longer periods, than those reported by
\citetalias{McQuillan_2014}.  A relevant flag for stars
missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.  Reviewing the resulting period detections, we
learned that 32 KOIs with orbital and rotation periods within
$\approx$20\% were excluded from the \citetalias{Santos_2019} and
\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
to include these objects.\footnote{In general, excluding stars in this
category would impose a bias against young close-in planets, due to
the natural commensurability between the rotation rates of young
stars, and Kepler's increased sensitivity to short-period objects.
While some of these KOIs are known false positives, our analysis
naturally excludes such objects at later steps.  As one motivating
example, this subset of stars happens to include {\it
e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
based on membership in RSG-5 \citep{Bouma_2022b}.  }
The final tally of stars from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our corrected KOI list is
\nuniqstarsantosrot\ Kepler stars with rotation periods.

@@

\subsection{Star Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
stars with reported rotation periods and effective temperatures.  To
select the subset stars for which we expect these age to in fact be
valid, we then build a set of quality flags.  When and
how these flags should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, they might
all be applied.
However if the goal is constructing a complete sample of young stars,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  
The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
selecting for such cases using bitflags.

@@

{\it Final calibration sample}---{\bf todo: mention any extra flags
from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
amplitude cuts you might want to apply to the rotation period catalog}

\subsection{Planet Quality Flags}
For the subset of our analysis that focuses on the properties of
planets, we use the following additional quality flags.

@@

The rotation periods for the stellar and planetary samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods yielded
\nuniqstarsantosrot.  Further requiring these rotating stars
to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
over which \texttt{gyro-interp} returns finite age posteriors,
yields \nuniqstarsantosrotteffcut\ stars.
Further requiring them to be apparently single, near the main
sequence, with $\log g$$>$4.2, and not flagged as
eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
potentially amenable for gyrochronology.
Finally, imposing RUWE$<$1.4 leaves \nuniqstarsantosrotgyroappl\ stars
for which gyrochronology is likely to be valid.

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.
Then,
require these confirmed and candidate planets to have MES$>$10.
This combination of selection criteria
yield \nplwgyroagewithgrazingandhighruwe\ planets with rotation-based
ages orbiting \nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$)
based on the Markov Chain Monte Carlo (MCMC) fits from
\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ and
\nplhoststarwgyroagenograzing\ planets and stars, respectively.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_eagles_showdispersion.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    from Keck/HIRES spectra collected between 2009 and 2024.}
    Lines represent mean models derived from Gaia-ESO spectra of stars
    in clusters of various ages; the intrinsic dispersion 
    evolves in time, and is modelled as the $\pm$1$\sigma$
    bands~\citep{Jeffries_2023}.
    As for rotation-based ages, the intrisic dispersion sets the
    empirical precision limit for the technique.
		\label{fig:li_vs_teff}
	}
\end{figure}

%TODO: does JUMP mirror all HIRES data from KOA?
We queried a combination of the Keck Observatory Archive and the
internal CPS archive to collect all HIRES spectra that were available
for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
with plausible rotation-based ages.
This exercise yielded at least one spectrum for \nlithiumstars\ and 
\nlithiumplanets\ planets.

Out of the \nplhoststarwgyroagewithgrazingandhighruwe\ stars for w

To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
with HIRES spectra, we use the procedure described by
\citet{Bouma_2021}.
The stars of interest for this work are FGK stars, and so the
continuum in the vicinity of the
\ion{Li}{1} 6708\,\AA\ doublet is well-defined.
We therefore numerically integrate a single best-fit Gaussian over a
local window, and estimate the uncertainties on the line width through
a Monte Carlo procedure that bootstraps against the local scatter in the
continuum.
Generally speaking, this approach does not correct for the neighboring
\ion{Fe}{1} 6707.44\,\AA\ blend.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Age distribution of stars and planet-hosts in the Kepler field
    inferred from rotation periods}.
    A histogram of samples from the age posteriors for all Kepler
    stars with reported rotation periods and temperatures is shown in
    light gray;
    the age calculation was actually valid for stars that meet the criteria
    discussed in Section~\ref{subsec:flags} (dark blue).
		\label{fig:hist_tgyro}
	}
\end{figure*}
